---
title: "phw251_medicare"
format: html
editor: visual
---

Scenario: Mortality and Medicare services in the United States

This scenario will utilize state-level mortality data for the United States (via CDC Wonder), population data by state and age group, and state-level Medicare enrollment and spending (via Center for Medicare Services and Kaiser Family Foundation). Teams will be asked to aggregate mortality data for a chosen category of cause of death and age group, calculate mortality rate for the specified cause and age group, and compare with a population-adjusted measure of Medicare utilization/enrollment.


Description: You work for a non-profit organization that is focused on making recommendations for public health related policy changes. Your organization is working on defending a policy proposal that would increase funding to Medicare and Medicaid programs.

Research Question: You have been asked to explore existing public data sources to determine if there is evidence of a correlation between mortality rates and health care expenditures in the United States. Additionally, you have been asked to highlight a few states that have higher mortality rates and lower public healthcare spending than other states.

```{r, echo=FALSE, include=FALSE, warning=FALSE}
library(tidyverse)
library(readr)
library(stringr)
```

```{r,  echo=FALSE, include=FALSE, warning=FALSE}
#Milestone 2: Import data
mort <- read_csv("data/cdc_wonder_mortality.csv")
pop <- read_csv("data/cdc_wonder_population.csv")
medicaid <-read_csv("data/kff_cms_medicaid_state_summary.csv")
medicare <-read_csv("data/kff_cms_medicare_state_summary.csv")
```

```{r, echo=FALSE, include=FALSE, warning=FALSE}
#Milestone 2: Identify key data elements and their data type
summary(mort)
#Residence State: Ten-Year Age Groups Code are all character data type
#Deaths is a double/numeric
#Range for Deaths (number of) is 10 to 16327
unique(mort$`Residence State`) #51 states + NA. REMOVE NA
unique(mort$`UCD - ICD Sub-Chapter`) #119 causes of mortality
unique(mort$`Ten-Year Age Groups`) #11 groups + NA

summary(pop)
#Range of population, 6738 to 5839642. Investigate 1 NA
#State and 10-yr age groups are both character
#State and state code need to be separated out

unique(pop$State) #51 + "NA". REMOVE NA
unique(pop$`Ten-Year Age Groups`) #11 + NA. Appears to be identical to groupings for mortality

summary(medicaid)
#Multiple columns with numeric values ARE READING AS CHARACTER because the values have $ in them.
#For these, if remember to clean any of those you will be using. Separate out the $ from the numeric and create a numeric value for these.
#Leave NAs and 0s in place for now
#Other than Local, all other values should be numeric

summary(medicare)
#a couple of columns saved as characters should be numeric - same issue as with medicaid, remove $ and convert to numeric. 

#Joining - can join all together by state name. Make sure all are named the same OR use state codes. Not all data sets have state codes. 
#RENAME so that all state and state code names match for easier joining

summary(medicare)
#TIME STAMP: 30 minutes
```

```{r}
#Milestone 2, task 2: clean each data set

#1. clean mortality data
mort_clean <- mort %>% 
  rename_with( ~ tolower(gsub(" ", "_", .x, fixed = TRUE)))%>%
  rename(ucd_icd_sub_chapter = `ucd_-_icd_sub-chapter`,
         ucd_icd_sub_chapter_code = `ucd_-_icd_sub-chapter_code`,
         age_groups = `ten-year_age_groups`,
         age_groups_code = `ten-year_age_groups_code`,
         state = residence_state,
         state_code = residence_state_code)

#2. clean population data
pop_clean <- pop %>%
  rename_with( ~ tolower(gsub(" ", "_", .x, fixed = TRUE)))%>%
  rename(age_groups = `ten-year_age_groups`,
         old_state = state)%>% 
  mutate(state_code = str_sub(old_state, start = 1, end = 2),
         state = str_sub(old_state, start = 4, end = -1))%>%
  select(-old_state)

#3. clean medicaid data
medicaid_clean <- medicaid %>%
  rename_with( ~ tolower(gsub(" ", "_", .x, fixed = TRUE))) %>%
  rename(state = location)%>%
  #remove $ signs
  mutate(state = str_to_title(state),
         seniors_spending = str_sub(seniors_spending, start = 2, end = -1),
         individuals_with_disabilities_spending = 
           str_sub(individuals_with_disabilities_spending, start = 2, end = -1),
         adult_spending = str_sub(adult_spending, start =2, end = -1),
         children_spending = str_sub(children_spending, start = 2, end = -1),
         newly_eligible_adult_spending = 
           str_sub(newly_eligible_adult_spending, start =2, end = -1),
         total_spending = str_sub(total_spending, start = 2, end = -1),
         #convert characters to numeric
         across(2:7, ~as.numeric(.x, na.rm = T)),
         newly_eligible_adult_enrollment = as.numeric(newly_eligible_adult_enrollment)) 

#4. clean medicare data  
medicare_clean <- medicare %>%
  rename_with( ~ tolower(gsub(" ", "_", .x, fixed = TRUE))) %>%
  rename(state = location)%>%
  mutate(state = str_to_title(state),
         #remove $$ signs
         `total_medicare_part_a_and/or_part_b_program_payments` =
           str_sub(`total_medicare_part_a_and/or_part_b_program_payments`, start = 2, end = -1),
         `total_medicare_part_a_program_payments` =
           str_sub(`total_medicare_part_a_program_payments`, start = 2, end = -1),
         `total_medicare_part_b_program_payments` =
           str_sub(`total_medicare_part_b_program_payments`, start = 2, end = -1),
         #remove commas
         `total_medicare_part_a_and/or_part_b_program_payments` =
           str_remove_all(`total_medicare_part_a_and/or_part_b_program_payments`, ","),
         `total_medicare_part_a_program_payments` =
           str_remove_all(`total_medicare_part_a_program_payments`, ","),
         `total_medicare_part_b_program_payments` =
           str_remove_all(`total_medicare_part_b_program_payments`, ",")
         )
    
```


```{r}
#Milestone 3
#Choose an age category and mortality grouping of interest - choosing ages 65-74 and "transport accidents" and "Other external causes of accidental injury"
#choose at least 2 ICD sub-chapters

#1. mortality data for ages 65-74, transport and accidental injuries
mort_65 <- mort_clean %>%
  filter(age_groups == "65-74 years",
         ucd_icd_sub_chapter_code == c("V01-V99", "W00-X59"))
#Create a new variable that includes a state level sum and summarize to only include one row per state
mort_accident <- mort_65 %>%
  group_by(state) %>% 
  summarize(accident_deaths = sum(deaths))

#2. population
pop_65 <- pop_clean %>% 
  #Create a new variable with population per state for the age group of interest and summarize to only include one row per state 
  filter(age_groups == "65-74 years")




#combine into a single state-level dataset
```

