---
title: "phw251_medicare"
format: html
editor: visual
---

Instructions: https://bcourses.berkeley.edu/courses/1525339/pages/team-project#S3

Scenario: Mortality and Medicare services in the United States

This scenario will utilize state-level mortality data for the United States (via CDC Wonder), population data by state and age group, and state-level Medicare enrollment and spending (via Center for Medicare Services and Kaiser Family Foundation). Teams will be asked to aggregate mortality data for a chosen category of cause of death and age group, calculate mortality rate for the specified cause and age group, and compare with a population-adjusted measure of Medicare utilization/enrollment.

Description: You work for a non-profit organization that is focused on making recommendations for public health related policy changes. Your organization is working on defending a policy proposal that would increase funding to Medicare and Medicaid programs.

Research Question: You have been asked to explore existing public data sources to determine if there is evidence of a correlation between mortality rates and health care expenditures in the United States. Additionally, you have been asked to highlight a few states that have higher mortality rates and lower public healthcare spending than other states.

```{r, echo=FALSE, include=FALSE, warning=FALSE}
library(tidyverse)
library(readr)
library(stringr)
library(plotly)
```

```{r,  echo=FALSE, include=FALSE, warning=FALSE}
#Milestone 2: Import data
mort <- read_csv("data/cdc_wonder_mortality.csv")
pop <- read_csv("data/cdc_wonder_population.csv")
medicaid <-read_csv("data/kff_cms_medicaid_state_summary.csv")
medicare <-read_csv("data/kff_cms_medicare_state_summary.csv")
```

```{r, echo=FALSE, include=FALSE, warning=FALSE}
#Milestone 2: Identify key data elements and their data type
summary(mort)
#Residence State: Ten-Year Age Groups Code are all character data type
#Deaths is a double/numeric
#Range for Deaths (number of) is 10 to 16327
unique(mort$`Residence State`) #51 states + NA. REMOVE NA
unique(mort$`UCD - ICD Sub-Chapter`) #119 causes of mortality
unique(mort$`Ten-Year Age Groups`) #11 groups + NA

summary(pop)
#Range of population, 6738 to 5839642. Investigate 1 NA
#State and 10-yr age groups are both character
#State and state code need to be separated out

unique(pop$State) #51 + "NA". REMOVE NA
unique(pop$`Ten-Year Age Groups`) #11 + NA. Appears to be identical to groupings for mortality

summary(medicaid)
#Multiple columns with numeric values ARE READING AS CHARACTER because the values have $ in them.
#For these, if remember to clean any of those you will be using. Separate out the $ from the numeric and create a numeric value for these.
#Leave NAs and 0s in place for now
#Other than Local, all other values should be numeric

summary(medicare)
#a couple of columns saved as characters should be numeric - same issue as with medicaid, remove $ and convert to numeric. 

#Joining - can join all together by state name. Make sure all are named the same OR use state codes. Not all data sets have state codes. 
#RENAME so that all state and state code names match for easier joining

summary(medicare)
#TIME STAMP: 30 minutes
```

```{r, echo=FALSE, include=FALSE, warning=FALSE}
#Milestone 2, task 2: clean each data set

#1. clean mortality data
mort_clean <- mort %>% 
  rename_with( ~ tolower(gsub(" ", "_", .x, fixed = TRUE)))%>%
  rename(ucd_icd_sub_chapter = `ucd_-_icd_sub-chapter`,
         ucd_icd_sub_chapter_code = `ucd_-_icd_sub-chapter_code`,
         age_groups = `ten-year_age_groups`,
         age_groups_code = `ten-year_age_groups_code`,
         state = residence_state,
         state_code = residence_state_code)

#2. clean population data
pop_clean <- pop %>%
  rename_with( ~ tolower(gsub(" ", "_", .x, fixed = TRUE)))%>%
  rename(age_groups = `ten-year_age_groups`,
         old_state = state)%>% 
  mutate(state_code = str_sub(old_state, start = 1, end = 2),
         state = str_sub(old_state, start = 4, end = -1))%>%
  select(-old_state)

#3. clean medicaid data
medicaid_clean <- medicaid %>%
  rename_with( ~ tolower(gsub(" ", "_", .x, fixed = TRUE))) %>%
  rename(state = location)%>%
  #remove $ signs
  mutate(state = str_to_title(state),
         seniors_spending = str_sub(seniors_spending, start = 2, end = -1),
         individuals_with_disabilities_spending = 
           str_sub(individuals_with_disabilities_spending, start = 2, end = -1),
         adult_spending = str_sub(adult_spending, start =2, end = -1),
         children_spending = str_sub(children_spending, start = 2, end = -1),
         newly_eligible_adult_spending = 
           str_sub(newly_eligible_adult_spending, start =2, end = -1),
         total_spending = str_sub(total_spending, start = 2, end = -1),
         #convert characters to numeric
         across(2:7, ~as.numeric(.x, na.rm = T)),
         newly_eligible_adult_enrollment = as.numeric(newly_eligible_adult_enrollment),
         ) 

medicaid_clean <- medicaid_clean %>%
  mutate(state = if_else(state == "District Of Columbia", "District of Columbia", state))

#4. clean medicare data  
medicare_clean <- medicare %>%
  rename_with( ~ tolower(gsub(" ", "_", .x, fixed = TRUE))) %>%
  rename(state = location)%>%
  mutate(state = str_to_title(state),
         #remove $$ signs
         `total_medicare_part_a_and/or_part_b_program_payments` =
           str_sub(`total_medicare_part_a_and/or_part_b_program_payments`, start = 2, end = -1),
         `total_medicare_part_a_program_payments` =
           str_sub(`total_medicare_part_a_program_payments`, start = 2, end = -1),
         `total_medicare_part_b_program_payments` =
           str_sub(`total_medicare_part_b_program_payments`, start = 2, end = -1),
         #remove commas
         `total_medicare_part_a_and/or_part_b_program_payments` =
           str_remove_all(`total_medicare_part_a_and/or_part_b_program_payments`, ","),
         `total_medicare_part_a_program_payments` =
           str_remove_all(`total_medicare_part_a_program_payments`, ","),
         `total_medicare_part_b_program_payments` =
           str_remove_all(`total_medicare_part_b_program_payments`, ",")
         )

medicare_clean <- medicare_clean %>%
  mutate(state = if_else(state == "District Of Columbia", "District of Columbia", state))    
```

```{r, echo=FALSE, include=FALSE, warning=FALSE}
#Milestone 3
#Choose an age category and mortality grouping of interest - choosing ages 65-74 and "transport accidents" and "Other external causes of accidental injury"
#choose at least 2 ICD sub-chapters

#1. mortality data for ages 65-74, transport and accidental injuries
mort_65 <- mort_clean %>%
  filter(age_groups %in% c("65-74 years", "75-84 years", "85+years"),
         ucd_icd_sub_chapter_code %in% c("V01-V99", "W00-X59")) %>% 
  select(state, ucd_icd_sub_chapter, ucd_icd_sub_chapter_code, age_groups, age_groups_code, deaths)
  
#Create a new variable that includes a state level sum and summarize to only include one row per state
mort_tot_deaths <- mort_65 %>%
  group_by(state) %>% 
  summarize(accident_deaths = sum(deaths))

#2. population
pop_65 <- pop_clean %>% 
  #Create a new variable with population per state for the age group of interest and summarize to only include one row per state 
  filter(age_groups == "65-74 years")
#there are 51 states

#3. medicaid
medicaid_65 <- medicaid_clean %>%
  select(state, seniors_spending, total_spending,
         seniors_enrollment, total_enrollment)%>%
  rename(seniors_medicaid_spending = seniors_spending,
         total_medicaid_spending = total_spending,
         seniors_medicaid_enrollment = seniors_enrollment,
         total_medicaid_enrollment = total_enrollment
         )%>%
  #Create a new variable for spending per enrollee for age group of interest
  mutate(sr_medicaid_spend_per_enrollee = round(seniors_medicaid_spending/seniors_medicaid_enrollment,2),
         tot_medicaid_spend_per_enrollee = total_medicaid_spending/total_medicaid_enrollment)
```

```{r, include=FALSE, echo=FALSE, warning=FALSE}
#there are 51 states
#4. medicare
#traditional vs. part a & b: traditional is part a only (requires no monthly premium). people that enroll in medicare are automatically enrolled in traditional. Part b requires monthly premium and you have to elect to accept or decline this coverage. You have to be enrolled in medicare parts A and B to be eligible to enroll in Medicare Advantage (data not included here). With MA you are choosing to have your Medicare benefit provided through a private plan.
#part A covers hospital care (hospital care, hospice, home health care, and skilled nursing facility)
#part B covers medical insurance (doctors visits, outpatient procedures, etc.)
medicare_65 <- medicare_clean %>%
  #I'm going to choose to look at just overall enrollment, utilization and payments at this point. 
  select(1:4)%>%
  rename(total_medicare_enrollees = `total_traditional_medicare_part_a_and/or_part_b_enrollees`,
         total_medicare_utilization = `total_medicare_part_a_and/or_part_b_persons_with_utilization`,
         total_medicare_payments = `total_medicare_part_a_and/or_part_b_program_payments`)%>%
  mutate(total_medicare_payments = as.numeric(total_medicare_payments),
         #Create a new variable for spending per enrollee for age group of interest
         medicare_spend_per_enrollee = total_medicare_payments/total_medicare_enrollees)
#all states present 

#calculate the total population for each state
tot_pop <- pop_clean %>% 
  group_by(state) %>% 
  summarize(tot_pop = sum(population))
```

```{r}
#combine into a single state-level dataset
df_pop_mort <- mort_65 %>%
  left_join(pop_65, by = c("state", "age_groups")) %>% 
  select(-c("age_groups_code", "state_code", ucd_icd_sub_chapter_code)) %>% 
  pivot_wider(names_from = c(age_groups, ucd_icd_sub_chapter), values_from = deaths) 

df_pop_mort[is.na(df_pop_mort)] <- 0

df_pop_mort <- df_pop_mort %>% 
  group_by(state) %>% 
  summarize(pop = sum(population),
            #I did these below to get data into shape to join up with the other data set
            transport_65_74 = sum(`65-74 years_Transport accidents`),
            transport_75_84 = sum(`75-84 years_Transport accidents`),
            other_acc_65_74 = sum(`65-74 years_Other external causes of accidental injury`),
            other_acc_75_84 = sum(`75-84 years_Other external causes of accidental injury`))

df_medicaid_care <- medicaid_65 %>% 
  left_join(medicare_65, by = "state")

df <- df_pop_mort %>% 
  full_join(df_medicaid_care, by = "state")%>% 
  group_by(state) %>% 
  left_join(tot_pop, by = "state")%>%
  rename(senior_pop = pop) %>% 
  mutate(
    tot_deaths = sum(transport_65_74+transport_75_84+other_acc_65_74+other_acc_75_84),
    #create a death rate per 100,000. I checked online and the calculated rate is not far off the rate listed by CDC, considering that this includes all seniors (rate goes up dramatically for older seniors)
    death_rate = (tot_deaths/senior_pop)*100000,
    death_rate = round(death_rate,2)
                       )%>% 
  select(state, 
         senior_pop, tot_pop, 
         transport_65_74, transport_75_84, 
         other_acc_65_74, other_acc_75_84,
         tot_deaths,
         death_rate,
         sr_medicaid_spend_per_enrollee, tot_medicaid_spend_per_enrollee,
         seniors_medicaid_spending, seniors_medicaid_enrollment,
         total_medicaid_spending, total_medicaid_enrollment
         )

df <- df %>% 
  filter(!state %in% 'United States')

```

```{r}
#correlation: create visualizations to show if there is correlation between total mortality rate and health care expenditures
#highlight a few states with the highest mortality rates and lowest public healthcare spending

ggplot(data = df, aes(x=death_rate, y = sr_medicaid_spend_per_enrollee))+
 geom_boxplot()
  
ggplot(data = df, aes(x=death_rate, y = sr_medicaid_spend_per_enrollee))+
  geom_point()+
  geom_smooth()

library(stats)
cor(x = df$death_rate, y = df$sr_medicaid_spend_per_enrollee, use = "complete.obs") #-0.0798

#there doesn't appear to be a strong correlation. However, there are several outliers with much fewer deaths and much higher spending
```

```{r}
library(plotly)

p1 <- ggplot(data = df, aes(x=death_rate, y = sr_medicaid_spend_per_enrollee, fill=state))+
  geom_point()+
  geom_smooth()

ggplotly(p1)

#we could see what the correlation looks like if you remove North Dakota, Pennsylvania, and Minnesota
```

```{r}
df_trimmed <- df %>%
  filter(!state %in% c('North Dakota', 'Pennsylvania', 'Minnesota', 'United States'))

p2 <- ggplot(data = df_trimmed, aes(x=death_rate, y = sr_medicaid_spend_per_enrollee, fill=state))+
  geom_point()+
  geom_smooth()

ggplotly(p2)

cor(x = df_trimmed$death_rate, y = df_trimmed$sr_medicaid_spend_per_enrollee, use = "complete.obs") #-0.0468. Even less of a correlation
```

```{r}
#make a table. give each state an percentile indicator for death rate and for spending. pinpoint states with highest mortality rates and lowest spending
library(kableExtra)
df_table <- df %>% 
  select(state, senior_pop, death_rate, sr_medicaid_spend_per_enrollee) 
#using ntile to rank the 51 states & DC
df_table$death_rank <- ntile(df_table$death_rate, 51)
#reverse rank on spending to show states with lowest spend
df_table$spend_rank <- ntile(-df_table$sr_medicaid_spend_per_enrollee, 51)
df_table$comb_rank <- df_table$death_rank + df_table$spend_rank

df_table <- df_table %>% 
  arrange(desc(comb_rank)) %>% 
  select(state, death_rank, spend_rank, comb_rank)

kbl(df_table)

#highlight a few states with the highest mortality rates and lowest public healthcare spending
#states with a combined rank of high death and low spend: Kentucky, Wisconsin, Nevada, Mississippi, and New Mexico
#states with a combined rank of lower death and higher spend: New York, New Jersey, North Dakota (they spend a lot), Massachusetts, and New Hampshire. North Dakota and Massachusetts were 2 of the outliers I noticed in my correlation plot.
```

```{r}
#next: create a searchable table
#create interactive boxplot and scatterplot
#think about other visuals to tell the story.
#what is the story here?
```

