---
title: "phw251_medicare"
format: html
editor: visual
---

Scenario: Mortality and Medicare services in the United States

This scenario will utilize state-level mortality data for the United States (via CDC Wonder), population data by state and age group, and state-level Medicare enrollment and spending (via Center for Medicare Services and Kaiser Family Foundation). Teams will be asked to aggregate mortality data for a chosen category of cause of death and age group, calculate mortality rate for the specified cause and age group, and compare with a population-adjusted measure of Medicare utilization/enrollment.


Description: You work for a non-profit organization that is focused on making recommendations for public health related policy changes. Your organization is working on defending a policy proposal that would increase funding to Medicare and Medicaid programs.

Research Question: You have been asked to explore existing public data sources to determine if there is evidence of a correlation between mortality rates and health care expenditures in the United States. Additionally, you have been asked to highlight a few states that have higher mortality rates and lower public healthcare spending than other states.

```{r, echo=FALSE, include=FALSE, warning=FALSE}
library(tidyverse)
library(readr)
```

```{r,  echo=FALSE, include=FALSE, warning=FALSE}
#Milestone 2: Import data
mortality <- read_csv("data/cdc_wonder_mortality.csv")
population <- read_csv("data/cdc_wonder_population.csv")
medicaid <-read_csv("data/kff_cms_medicaid_state_summary.csv")
medicare <-read_csv("data/kff_cms_medicare_state_summary.csv")
```

```{r}
#Milestone 2: Identify key data elements and their data type
summary(mortality)
#Residence State: Ten-Year Age Groups Code are all character data type
#Deaths is a double/numeric
#Range for Deaths (number of) is 10 to 16327
unique(mortality$`Residence State`) #51 states + NA. REMOVE NA
unique(mortality$`UCD - ICD Sub-Chapter`) #119 causes of mortality
unique(mortality$`Ten-Year Age Groups`) #11 groups + NA

summary(population)
#Range of population, 6738 to 5839642. Investigate 1 NA
#State and 10-yr age groups are both character
#State and state code need to be separated out

unique(population$State) #51 + "NA". REMOVE NA
unique(population$`Ten-Year Age Groups`) #11 + NA. Appears to be identical to groupings for mortality

summary(medicaid)
#Multiple columns with numeric values ARE READING AS CHARACTER because the values have $ in them.
#For these, if remember to clean any of those you will be using. Separate out the $ from the numeric and create a numeric value for these.
#Leave NAs and 0s in place for now
#Other than Local, all other values should be numeric

summary(medicare)
#a couple of columns saved as characters should be numeric - same issue as with medicaid, remove $ and convert to numeric. 


#Joining - can join all together by state name. Make sure all are named the same OR use state codes. Not all data sets have state codes. 
#RENAME so that all state and state code names match for easier joining

summary(medicare)
#TIME STAMP: 30 minutes
```

```{r}
#Milestone 2, task 2: clean each data set

```


```{r}
#Milestone 3
#Choose an age category and mortality grouping of interest
```

